Hospitalizations for psychiatric conditions among adolescents during the COVID-19 pandemic

### 2021-10-01 psy_peds_3.Rmd
obfuscation added and time period as a variable

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "../Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
data_update_date <- "2021-06-05"
```

@@

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))
```

@@

```{r}

new_hospitalizations_per_week <- clinical_table %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(weeks)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num)) %>% 
  ungroup()

@@

nh_week <- admission_plot_count(df = new_hospitalizations_per_week, 
                     x = "weeks",
                     y = "count", 
                     title = "New admissions (per week)", 
                     y_title = "Patient Count")

nh_week
```

@@

@@

```{r}
unique_patients_hospitalized_per_week <- clinical_table_ado %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num))

th_week <- admission_plot_count(df = unique_patients_hospitalized_per_week, 
                     x = "weeks",
                     y = "count", 
                     title = "Total hospitalization (per week)", 
                     y_title = "Patient Count") +
  labs( subtitle = "(age range 11-17)")
th_week
```

@@

all_codes_desc <- all_codes_ado %>% 
    left_join( icdCodes, by=c("concept_code" = "ICD10_Code") ) %>%
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "others")) %>%
  dplyr::select( patient_num, sex, age_time_diagnosis, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, month, year,
                 period, disorder_group, description, icd_code_category )

@@

```{r}
count_icd_per_week_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count_psy = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd_per_week <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count_icd = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd_per_week <- count_icd_per_week %>%
  left_join(count_icd_per_week_psy, 
            by = c("weeks", "period")) %>% 
  replace_na(list(count_psy = 0))

count_icd_per_week$percentage_psy <- 100*count_icd_per_week$count_psy / count_icd_per_week$count_icd

thpd_psyCount <- count_icd_per_week %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = weeks, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Date (by week)",
       title = "Weekly patient counts with mental health related ICD codes") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

```{r}
thpd_psyPer <- count_icd_per_week %>%
  ggplot(aes(x = weeks, y = percentage_psy, fill=period)) +
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  labs(y = "Percentage",
       x = "Date (by week)",
       title = "Percentage of patients with mental health related ICD codes (weekly)") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)+
  geom_point()
thpd_psyPer
```

@@

##### Total hospitalizations Per week

```{r}
count_disorder_group_per_week <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks, disorder_group ) %>%
  summarise(count_dg = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

dg_per_week <- count_disorder_group_per_week %>%  
  ggplot(aes(x = weeks, y= count_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Weeks",
       y = "Patient Count",
       title = "Weekly patient counts with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dg_per_week
```

@@

```{r}
# estimate the weekly counts
count_icd_per_week <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count_icd = n_distinct(patient_num)) %>%
  mutate( period = ifelse( weeks < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

# join the total with the disorder group count to estimate the percentage
perc_disorder_group_per_week <- count_icd_per_week %>%
  left_join(count_disorder_group_per_week, 
            by = c("weeks", "period")) %>% 
  replace_na(list(count_dg = 0))

perc_disorder_group_per_week$percentage_dg <- 100*perc_disorder_group_per_week$count_dg / perc_disorder_group_per_week$count_icd

dgPerc_per_week <- perc_disorder_group_per_week %>%  
  ggplot(aes(x = weeks, y= percentage_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = "Weeks",
       y = "Patient (%)",
       title = "Percentage of weekly patients with mental health related ICD codes", 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dgPerc_per_week
```

@@

```{r}
# counts per week and sex
patients_hospitalized_per_week_agg_sex <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( weeks, sex ) %>%
  summarise(count_sex = n_distinct(patient_num))

patients_hospitalized_per_week_agg_sex %>%
  ggplot(aes(x = weeks, y = count_sex, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = "Week",
       title = "Weekly patient counts aggregated by sex",
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)

##percentage
total_patients_psy_per_week <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( weeks ) %>%
  summarise(count = n_distinct(patient_num))

patients_hospitalized_per_week_agg_sex_perc <- total_patients_psy_per_week %>%
  left_join( patients_hospitalized_per_week_agg_sex, by = c("weeks")) %>% 
  replace_na(list(count_sex = 0))

patients_hospitalized_per_week_agg_sex_perc$percentage <- 100*patients_hospitalized_per_week_agg_sex_perc$count_sex/patients_hospitalized_per_week_agg_sex_perc$count

patients_hospitalized_per_week_agg_sex_perc %>%
  ggplot(aes(x = weeks, y = percentage, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Percentage ",
       x = "Week",
       title = "Weekly patient (%) aggregated by sex",
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```

>>>>

```{r message=FALSE, warning=FALSE}
folder_4ce_files <- "../Input/09212021_2/"
obfuscation =  FALSE
#obfuscation = 3
raceAvailable = TRUE
#raceAvailable = FALSE
dateFormat <- "%d-%b-%y"
min_hosp_days <- 2
data_update_date <- "2021-06-05"
time_period <-"weeks" #possible values: days, weeks, months
```

@@

clinical_filtered <- clinical_raw %>%
  filter( len_hospitalisation > min_hosp_days, 
          cohort  %in% c("AllAdm")) %>%
  dplyr::mutate( date = as.Date( calendar_date, format = dateFormat ),
                 weeks = as.Date(cut( date, breaks = "week")),
                 month = as.Date(cut( date, breaks = "month")),
                 year = format( date, "%Y"), 
                 period = ifelse( date < pandemic_start_date,
                                  "before_pandemic", "during_pandemic"))

clinical_filtered$time_p <- NA
if( time_period == "weeks"){
  clinical_filtered$time_p <- clinical_filtered$weeks
}else if(time_period == "days"){
  clinical_filtered$time_p <- clinical_filtered$date
}else if(time_period == "months"){
  clinical_filtered$time_p <- clinical_filtered$month
  }
```

@@

```{r}
new_hospitalizations <- clinical_table %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(time_p)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>% 
  ungroup()

@@

nh <- admission_plot_count(df = new_hospitalizations, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("New admissions (per ", time_period,")"), 
                     y_title = "Patient Count")

nh
```

@@

### New admissions age range 11-17

```{r}
new_hospitalizations_ado <- clinical_table_ado %>%
  group_by(patient_num, n_hospitalisation) %>%
  slice(which.min(time_p)) %>%
  filter( in_hospital == 1 ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>% 
  ungroup()

nh_ado <- admission_plot_count(df = new_hospitalizations_ado, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("New admissions (per ", time_period, ")"), 
                     y_title = "Patient Count")+
    labs( subtitle = "(age range 11-17)")

nh_ado
```

@@

```{r}
unique_patients_hospitalized <- clinical_table_ado %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count = ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

th <- admission_plot_count(df = unique_patients_hospitalized, 
                     x = "time_p",
                     y = "count", 
                     title = paste0("Total hospitalization (per ", time_period, ")"), 
                     y_title = "Patient Count") +
  labs( subtitle = "(age range 11-17)")
th
```

@@

all_codes_desc <- all_codes_ado %>% 
    left_join( icdCodes, by=c("concept_code" = "ICD10_Code") ) %>%
  mutate(icd_code_category = ifelse(concept_code %in% icdCodes$ICD10_Code, 
                               "psy", 
                               "others")) %>%
  dplyr::select( patient_num, sex, age_time_diagnosis, concept_type, concept_code, calendar_date,
                 in_hospital, n_hospitalisation, len_hospitalisation, date, weeks, month, year,
                 period, time_p, disorder_group, description, icd_code_category )

@@

```{r}
count_icd_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_psy =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_icd =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

count_icd <- count_icd %>%
  left_join(count_icd_psy, 
            by = c("time_p", "period")) %>% 
  replace_na(list(count_psy = 0))

count_icd$percentage_psy <- 100*count_icd$count_psy / count_icd$count_icd

thpd_psyCount <- count_icd %>%
  pivot_longer(names_to = "counts", cols=c(count_psy, count_icd)) %>%
  mutate( counts = ifelse( counts == "count_psy", "Psy", "Total")) %>%
  ggplot(aes(x = time_p, y = value, fill = counts, color = counts)) +
  geom_point() +
  # geom_line() +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  scale_fill_manual(values = cbPalette) +
  scale_color_manual(values = cbPalette) +
  labs(y = "Counts",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Per ", time_period , " patient counts with mental health related ICD codes")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)
thpd_psyCount

@@

```{r}
thpd_psyPer <- count_icd %>%
  ggplot(aes(x = time_p, y = percentage_psy, fill=period)) +
  scale_fill_manual(values = cbPalette[c(3,4)]) +
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  labs(y = "Percentage",
       x = paste0("Date (by ", time_period,")"),
       title = paste0("Percentage of patients with mental health related ICD codes (per ", time_period,")")) + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=period), colour= "black", size = 0.5)+
  geom_point()
thpd_psyPer
```

@@

##### Total hospitalizations

```{r}
count_disorder_group <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p, disorder_group ) %>%
  summarise(count_dg =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

dg <- count_disorder_group %>%  
  ggplot(aes(x = time_p, y= count_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient Count",
       title = paste0("Per ", time_period, " patient counts with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dg
```

@@

```{r}
# estimate the counts
count_icd <- all_codes_desc %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count_icd =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5)) %>%
  mutate( period = ifelse( time_p < pandemic_start_date,
                           "before_pandemic",
                           "during_pandemic")
  )

# join the total with the disorder group count to estimate the percentage
perc_disorder_group <- count_icd %>%
  left_join(count_disorder_group, 
            by = c("time_p", "period")) %>% 
  replace_na(list(count_dg = 0))

perc_disorder_group$percentage_dg <- 100*perc_disorder_group$count_dg / perc_disorder_group$count_icd

dgPerc <- perc_disorder_group %>%  
  ggplot(aes(x = time_p, y= percentage_dg, fill = period, group = disorder_group )) +
  geom_bar(stat = "identity")+
  geom_vline(xintercept = as.Date(pandemic_start_date),
             linetype = "dashed") +
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se= TRUE, aes(colour=period), colour= "black", size = 0.5) + 
  labs(x = time_period,
       y = "Patient (%)",
       title = paste0("Percentage of patients per ", time_period,"with mental health related ICD codes"), 
       subtitle = "Grouped by mental health ICD subcategories") + 
  ylim(0, NA) + 
  facet_wrap(. ~ disorder_group, scales = "free_y") +
  theme(strip.text.x = element_text(size = 4),
        axis.text.x = element_text(size = 5, angle = 90), 
        axis.text.y = element_text(size = 6), 
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10),
        title = element_text(size = 12),
        plot.subtitle = element_text(size = 11),
        legend.position = "none") +
  scale_fill_manual(values = cbPalette[-1])

dgPerc
```

@@

```{r}
# counts per time period and sex
patients_hospitalized_agg_sex <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots & 
           sex %in% c("female", "male")) %>%
    group_by( time_p, sex ) %>%
  summarise(count_sex =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

patients_hospitalized_agg_sex %>%
  ggplot(aes(x = time_p, y = count_sex, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Counts",
       x = time_period,
       title = paste0("Patient counts per ", time_period, " aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)

##percentage
total_patients_psy <- all_codes_psy %>%
  filter(date < end_date_plots &
         date >= start_date_plots ) %>%
    group_by( time_p ) %>%
  summarise(count =  ifelse(n_distinct(patient_num) > obfuscation | isFALSE(obfuscation),
                               n_distinct(patient_num), 
                               0.5))

patients_hospitalized_agg_sex_perc <- total_patients_psy %>%
  left_join( patients_hospitalized_agg_sex, by = c("time_p")) %>% 
  replace_na(list(count_sex = 0))

patients_hospitalized_agg_sex_perc$percentage <- 100*patients_hospitalized_agg_sex_perc$count_sex/patients_hospitalized_agg_sex_perc$count

patients_hospitalized_agg_sex_perc %>%
  ggplot(aes(x = time_p, y = percentage, fill = sex)) +
  geom_bar(stat="identity", position = "dodge")+
  scale_fill_manual(values = cbPalette) +
  labs(y = "Percentage ",
       x = time_period,
       title = paste0("Patient (%) per ", time_period," aggregated by sex"),
       subtitle = "Per Sex") + 
  theme(legend.position = "bottom") + 
  geom_smooth(method = "lm", formula = y ~ splines::ns(x, 2), se = TRUE, aes(colour=sex), colour= "black", size = 0.5)
```
